---
title: "Midi manipulation with pyramidi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midi manipulation with pyramidi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After installing pyramidi, you can run the helper function to install its python dependency miditapyr 
in your environment (in this case `"r-reticulate"`):

```{r}
pyramidi::install_miditapyr(envname = "r-reticulate")
```

## Load libraries

When hopefully everything is set up correctly, we'll load some libraries.

```{r setup, message=FALSE}
library(pyramidi)
library(dplyr)
library(purrr)
library(htmltools)
library(tibble)
library(details)
# This is a small helper package, that allows to synthesize & play midi files:
library(raudiomate) # see https://github.com/urswilke/raudiomate
```

## `MidiFramer` class

This class is the main structure of the package.

### Generate `MidiFramer` object

You can create an empty R6 object of class `"MidiFramer"` like so:

```{r}
mfr <- MidiFramer$new()
```


```{details, details.summary = "Show print output"}
mfr
```



But in this vignette, we'll create one from a `midi_file_path` by passing it to the  
constructor method `MidiFramer$new()`: 

```{r}
midi_file_path <- system.file("extdata", "test_midi_file.mid", package = "pyramidi")
mfr <- MidiFramer$new(midi_file_path)
class(mfr)
```

### Fields in `MidiFramer` objects

It is an [R6](https://r6.r-lib.org/) object that contains the following fields:

```{details, details.summary = "Show print output"}
enframe(as.list(mfr))
```

Please refer to `help(MidiFramer)` for more information on this class. The field `mf` is a [`miditapyr.MidiFrames()`](https://miditapyr.readthedocs.io/en/latest/miditapyr.html#miditapyr.midi_frame.MidiFrames) object. After its first element `mfr$mf$midi_file`, a [mido midi file object](https://mido.readthedocs.io/en/latest/midi_files.html) with the mido message data, there are 3 more dataframes:

```{details, details.summary = "Show print output of 30 first rows of these dataframes"}
list(
    mfr$mf$midi_frame_raw,
    mfr$mf$midi_frame_unnested$df,
    mfr$mf$midi_frame_nested$df
  ) %>% 
  map(head, 30) %>% 
  map(knitr::kable)
```

When `mf` is initialized `midi_frame_raw` and `midi_frame_nested$df` should be the same (except the ordering of the named fields in the `msg` column might differ). 

For a detailed interactive overview with further links showing how the various fields in the `MidiFramer` class are related and calculated see `vignette("package_workflow")`.

## Usage

### Modifying midi data

In the `MidiFramer` object, we can modify `mfr$df_notes_wide`, the notes in notewise wide format
(`note_on` & `note_off` events in the same line). Thus we don't need to worry which 
midi events belong together

Let's look at a small example. We'll define a function to replace every 
note with a random midi note between 60 & 71:

```{r}
mod <- function(dfn, seed) {
  n_notes <- sum(!is.na(dfn$note))
  dfn %>% mutate(note = ifelse(
    !is.na(note),
    sample(60:71, n_notes, TRUE),
    note
  ))
}
```

We could modify the notes in wide format like this:

```{r}
mod(mfr$df_notes_wide)
```

Then we would have to adapt all the following elements of `mfr` that depend on `mfr$df_notes_wide`.

When we call the method `mfr$update_notes_wide()`, all the depending list elements are also automatically updated.

```{r, message=FALSE}
# Apply the modification to mfr$df_notes_wide and all depending dataframes:
mfr$update_notes_wide(mod)
```

The data has also been changed in `mfr$mf` the miditapyr `MidiFrames` object in mfr:

```{r}
mfr$mf$midi_frame_nested$df %>% head()
```


### Writing modified midi files

Thus we can now directly save the modifications to a midi file:

```{r}
mfr$mf$write_file("mod_test_midi_file.mid")
```


### Playing audio

**You need to install fluidsynth if you want to do that on your computer.**
**To generate mp3 files, you also need ffmpeg.**

If you want to produce audio files from the midi files you can synthesize them
with the convenience function `synthesize_midi()`

```{r}
midifile <- "mod_test_midi_file.mid"
mp3file <- "mod_test_midi_file.mp3"
wavfile <- "mod_test_midi_file.wav"
```


```{r, eval=FALSE}
synthesize_midi(midifile, wavfile)
```

You can also listen to your synthesized files by embedding an audio player for the 
`MidiFramer` object:

```{r, eval=FALSE}
mfr$play(mp3file)
```

```{r echo=FALSE}
htmltools::tags$audio(
    controls = "",
    htmltools::tags$source(
      src = mp3file,
      type = "audio/mp3"
    )
  )
```


To play the audio, this saves the synthesized file under the path specified in `mp3file`. 
As you can hear, the `sample()` function in `mod()` changed all midi notes randomly.

### Multiple results

You can also generate a list of multiple `MidiFramer` objects (in this case 2) and apply different modifications to each:


```{r, eval=FALSE}
l_mfr <- 1:2 %>%
  set_names(paste0("test", ., ".mp3")) %>% 
  map(~mfr$clone(deep = TRUE)$update_notes_wide(mod))
```

And if you want to successively add modifications to the same object and store all intermediate results in a list, you could do it like this:

```{r, eval=FALSE}
l_mfr2 <- 1:2 %>%
  set_names(paste0("test", ., ".mp3")) %>% 
  accumulate(~.x$clone(deep = TRUE)$update_notes_wide(mod), .init = mfr)
```

Please note that we made deep copies (`$clone(deep = TRUE)`) of the `MidiFramer` object in the list `l_mfr`.
Otherwise the field of the python [`miditapyr.MidiFrames`](https://miditapyr.readthedocs.io/en/latest/miditapyr.html#miditapyr.midi_frame.MidiFrames) object `mf` is a shallow copy. 
This means that updates to one of these elements in `l_mfr` (for instance `l_mfr[[1]]$mf`) also change the others, because they all are references to the same object.


This is how you can embed multiple audio files in one call using `purrr::imap()`:

```{r, eval=FALSE}
tagList(
  imap(
    l_mfr, 
    ~ div(
      h4(paste("audio result", .y)),
      .x$play(audiofile = .y),
      # add 2 line breaks to vertically separate a bit:
      br(), 
      br()
    )
  )
)
```

```{r, echo=FALSE}
tagList(list(
  div(
    h4("audio result test1.mp3"),
    htmltools::tags$audio(
      controls = "",
      htmltools::tags$source(
        src = "test1.mp3",
        type = "audio/mp3"
      )
    ),
    br(),
    br()
  ),
  div(
    h4("audio result test2.mp3"),
    htmltools::tags$audio(
    controls = "",
    htmltools::tags$source(
      src = "test2.mp3",
      type = "audio/mp3"
    )
  )

  )
))
```

