---
title: "Using mido directly from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mido directly from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

By using `install_miditapyr()`, this should also install [mido](https://github.com/mido/mido) which is a 
dependency of miditapyr:

```{r, message=FALSE}
pyramidi::install_miditapyr(envname = "r-reticulate")
```

Now we'll load some R and the python mido & miditapyr libraries with the `reticulate::import()` function.
The loading of `import_builtins()` into `pb` will allow to work with python objects from R:


```{r setup}
library(reticulate)
library(pyramidi)
library(tidyr)
library(purrr)
library(details)
mido <- import("mido")
mt <- import("miditapyr")
pb <- import_builtins()
```

## Load file

Here we'll load the midi file included in the package into a  [`mido.MidiFile`](https://mido.readthedocs.io/en/latest/lib.html#mido.MidiFile) object:

```{r}
midi_file_string <- system.file("extdata", "test_midi_file.mid", package = "pyramidi")
midi_file <- mido$MidiFile(midi_file_string)
```


```{details, details.summary = "Click here to show the printed mido `MidiFile()` object"}
midi_file
```

This object is a list object, containing besides some meta information
multiple `MidiTrack()`s which are themselves also lists containing the midi event
messages.

## Transform to unnested dataframe 

We can transform the midi data to a data.frame similar to the one returned by
[`unnest_midi()`](https://miditapyr.readthedocs.io/en/latest/miditapyr.html#miditapyr.mido_io.unnest_midi)
from the miditapyr package. We'll only use the data of the first track of the midi file:


```{r}
# access first track:
track1 <- midi_file$tracks[[1]]
# transform MidiTrack() object to an R list of mido Message()s and MetaMessage()s:
track1list <- pb$list(track1)
```


```{details, details.summary = "Show details of list in R"}
track1list
```

The list contains mido `Message()` & `MetaMessage()` objects (`Message()`s seem to have a special print method).

With the function `pb$vars()` we can extract all variables for every 
event into a list of named R lists:

```{r}
track1Rlist <- track1list %>% 
  map(pb$vars)
```


```{details, details.summary = "Show details of first 10 elements of track events list in R"}
track1Rlist[1:10]

```

We can put this list of message lists in a dataframe and write the values of 
each key in their own column:


```{r}
track1Rlist %>% 
  tibble(temp = .) %>% 
  unnest_wider(temp)
```


## Compare with pyramidi

For comparison, we plot the results of [`unnest_midi()`](https://miditapyr.readthedocs.io/en/latest/miditapyr.html#miditapyr.mido_io.unnest_midi)


```{r}
mfr <- MidiFramer$new(midi_file_string)
mt$unnest_midi(dfc = mfr$mf$midi_frame_raw) %>% 
  as_tibble()
```

There are some differences in the types and `track1Rlist` only contains the data of track 1, but you can see that apart from that, the output is the same.
