---
title: "Package workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, echo=FALSE, results = 'hide'}
pyramidi::install_miditapyr(envname = "r-reticulate")
```

The following diagram depicts the structure of the pyramidi package and its
python counterpart [miditapyr](https://github.com/urswilke/miditapyr).

*Hover over nodes to see additional information. Many of the nodes are also 
clickable for further information.*

```{r setup, echo=FALSE}
library(pyramidi)
library(stringr)
library(purrr)
library(DiagrammeR)

mt_funs <- c("frame_midi", "unnest_midi", "nest_midi", "write_midi")

mt_fun_descrs <- map_chr(mt_funs, ~miditapyr[[.x]]$`__doc__`) %>% 
  str_split("\\n\\n") %>% 
  map_chr(1) %>% 
  str_replace_all("\\n", " ") %>% 
  str_squish() %>% 
  str_remove_all(":.*?:") %>% 
  str_remove_all("[`~]") 


pm_funs <- c(
  "pivot_long_notes",
  "pivot_wide_notes",
  "tab_measures",
  "merge_midi_frames",
  "split_midi_frame"
)
# inspired by:
# https://stackoverflow.com/a/8928894 &
# https://stackoverflow.com/a/23897916
get_help <- function(...){
    thefile <- help(...)
    capture.output(
        tools:::Rd2txt(utils:::.getHelpFile(thefile))
    )
}


pm_fun_descrs_raw <- map(pm_funs, get_help) %>% 
  map(~.x[-c(1:4)])


title_until_line_ints <- pm_fun_descrs_raw %>% 
  map_dbl(~str_which(.x, "^$") %>% min() %>% {. - 1})

pm_fun_descrs <- map2_chr(
  pm_fun_descrs_raw,
  title_until_line_ints,
  ~ .x[1:.y] %>% paste(collapse = " ") %>% str_squish() %>% str_remove_all("[`'\"’‘]")
)
gv <- grViz("package_workflow.gv")
```


```{r, echo=FALSE, fig.height = 8, fig.width = 8, out.width = "8.5in"}
# chunk options from here: https://stackoverflow.com/a/43213737
gv
```

* `miditapyr` can be used to load the midi file `midi_in` into a `mido.MidiFile`. 
The chain down to `midi_out` shows, how miditpayr's functions can be used 
to transform the data into different formats and finally save it to a new midi file. 
Have a look [here](https://miditapyr.readthedocs.io/en/latest/notebooks/functions_usage.html) 
to see in detail how this can be done.
* `mf` shows the structure of a
[`MidiFrames`](https://miditapyr.readthedocs.io/en/latest/miditapyr.html#miditapyr.midi_frame.MidiFrames) 
object and how its fields are related to the low-level functions. Have a look
[here](https://miditapyr.readthedocs.io/en/latest/notebooks/midi_frame_usage.html) 
to see how `MidiFrames` can be used in python.
* The `pyramidi` subgraph illustrates how the functions in this package work.
Starting from the dataframe `midi_frame_unnested`, the data is transformed to 
different formats, finally resulting in the same format.
* The dashed borders of the nodes in the graph depict the fields that are in a
`MidiFramer` object (for clarity not included in the graph). 
* The `MidiFramer` method `update_nodes_wide()` can be used to modify the 
field `df_notes_wide`. The red arrows and borders show the other fields of the 
`MidiFramer` object that are automatically updated when using this method.
