---
title: "Using the tuneR package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the tuneR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval=FALSE,
  comment = "#>"
)
```

This vignette shows how  a midi file can also be read in by the [tuneR](https://cran.r-project.org/package=tuneR) package and then written to 
disc. However, for other midi files the code might need to be adapted a bit.

### Load libraries

```{r setup}
library(pyramidi)
library(zeallot)
library(dplyr)
library(tidyr)
library(ggplot2)
```

<!-- ```{r, echo=FALSE} -->
<!-- # needed as long as the python modules of the package are not loaded correctly: -->
<!-- miditapyr <- reticulate::import("miditapyr") -->
<!-- mido <- reticulate::import("mido") -->

<!-- ``` -->

## Read midi in tidy format

### pyramidi

With pyramidi data can be loaded into a tidy dataframe like this:

```{r read_pyramidi}

mid_file_str <- system.file("extdata", "test_midi_file.mid", package = "pyramidi")
mido_mid_file <- mido$MidiFile(mid_file_str)
dfc <- miditapyr$midi_to_df(mido_mid_file)
ticks_per_beat = mido_mid_file$ticks_per_beat
df <- dfc %>%
  miditapyr$tidy_df() %>% 
  as_tibble()
df
```

### tuneR

With tuneR this could be done for this file in the following way:

```{r read}
# tuneR:
dft <- tuneR::readMidi(mid_file_str)
```

 (*for other midi files the cleaning code would have to be adapted*)



#### Changing format in dataframe

* to make compatible with pyramidi

```{r dataformat}
# vector according to which we'll
# rename the event types to the names given by mido, cf.:
# https://mido.readthedocs.io/en/latest/message_types.html &
# https://mido.readthedocs.io/en/latest/meta_message_types.html
rename_type_vec <-
  c(
    "Set Tempo"           = "set_tempo",
    "Time Signature"      = "time_signature",
    "Note On"             = "note_on",
    "Note Off"            = "note_off",
    "Sequence/Track Name" = "track_name",
    "End of Track"        = "end_of_track",
    "Key Signature"       = "key",
    "Controller"          = "control",
    "Program Change"      = "program",
    "Pitch Bend"          = "pitch"
  )


# transform the tuneR data.frame to the pyramidi format :
df2 <- dft %>%
  mutate(event = as.character(event)) %>%
  as_tibble() %>%
  mutate(meta = !is.na(type)) %>%
  select(-type) %>%
  rename(type     = event,
         note     = parameter1,
         velocity = parameter2,
         i_track  = track) %>%
  mutate(type  = recode(type, !!!rename_type_vec)) %>%
  mutate(tempo = ifelse(type == "set_tempo", as.integer(parameterMetaSystem), NA)) %>%
  mutate(name  = ifelse(type == "track_name", parameterMetaSystem, NA)) %>%
  mutate(temp  = ifelse(type == "time_signature", parameterMetaSystem, NA)) %>%
  separate(temp, c("numerator", 
                   "denominator", 
                   "clocks_per_click", 
                   "temp1", "temp2", 
                   "notated_32nd_notes_per_beat"),
           convert = TRUE) %>%
  select(-temp1, -temp2) %>%
  arrange(i_track, time) %>%
  group_by(i_track) %>%
  # calculate time increments instead of total time:
  mutate(time = time - lag(time) %>% {.[1] = 0; .}) %>%
  ungroup() %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), NaN, .)) %>%
  mutate(name = ifelse(is.na(name), list(NULL), name))

```

### Apply pyramidi functions


```{r }
dfm2 <- tab_measures(df2, ticks_per_beat) %>%
  # create a variable `track` with the track name (in order to have it in the
  # plot below)
  mutate(track = ifelse(purrr::map_chr(name, typeof) != "character", 
                        list(NA_character_), 
                        name)) %>%
  unnest(cols = track) %>% 
  fill(track)


dfm2 %>%
  miditapyr$split_df()  %->% c(df_meta, df_notes)

# small correction of transforming "NA" strings into real NA values:
df_meta <-
  df_meta %>%
  mutate(name = ifelse(name == "NA", NA, name))
```

Now we can do exactly the same as if this data was read in by pyramidi (cf. the [README](https://github.com/urswilke/pyramidi/#readme))

```{r child='../man/rmdhunks/piano_roll_note_date.Rmd'}
```

```{r child='../man/rmdhunks/midi_df_to_file.Rmd'}
```

