---
title: "Using the tuneR package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the tuneR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how a midi file can also be read in by the [tuneR](https://cran.r-project.org/package=tuneR) package and then written to 
disc. However, for other midi files the code might need to be adapted a bit.

## Load libraries

```{r setup, message=FALSE}
library(pyramidi)
library(zeallot)
library(dplyr)
library(tidyr)
library(ggplot2)
```


## Read midi in unnested format

### pyramidi

With pyramidi data can be loaded into a unnested dataframe like this:

```{r read_pyramidi}

mid_file_str <- system.file("extdata", "test_midi_file.mid", package = "pyramidi")
mido_mid_file <- mido$MidiFile(mid_file_str)
dfc <- miditapyr$frame_midi(mido_mid_file)
ticks_per_beat = mido_mid_file$ticks_per_beat
df <- dfc %>%
  miditapyr$unnest_midi() %>% 
  as_tibble()
df
```

### tuneR

With tuneR this can be done for this file in the following way:

```{r read}
df_tuner <- tuneR::readMidi(mid_file_str)
```

(*for other midi files the cleaning code would have to be adapted*)



### Changing format in tuneR dataframe

* to make compatible with pyramidi

```{r dataformat}
# vector according to which we'll
# rename the event types to the names given by mido, cf.:
# https://mido.readthedocs.io/en/latest/message_types.html &
# https://mido.readthedocs.io/en/latest/meta_message_types.html
rename_type_vec <-
  c(
    "Set Tempo"           = "set_tempo",
    "Time Signature"      = "time_signature",
    "Note On"             = "note_on",
    "Note Off"            = "note_off",
    "Sequence/Track Name" = "track_name",
    "End of Track"        = "end_of_track",
    "Key Signature"       = "key",
    "Controller"          = "control",
    "Program Change"      = "program",
    "Pitch Bend"          = "pitch"
  )


# transform the tuneR data.frame to the pyramidi format :
df_tuner_mod <- df_tuner %>%
  mutate(event = as.character(event)) %>%
  as_tibble() %>%
  mutate(meta = !is.na(type)) %>%
  select(-type) %>%
  rename(type     = event,
         note     = parameter1,
         velocity = parameter2,
         i_track  = track) %>%
  mutate(type  = recode(type, !!!rename_type_vec)) %>%
  mutate(tempo = ifelse(type == "set_tempo", as.integer(parameterMetaSystem), NA)) %>%
  mutate(name  = ifelse(type == "track_name", parameterMetaSystem, NA)) %>%
  mutate(temp  = ifelse(type == "time_signature", parameterMetaSystem, NA)) %>%
  separate(temp, c("numerator", 
                   "denominator", 
                   "clocks_per_click", 
                   "temp1", "temp2", 
                   "notated_32nd_notes_per_beat"),
           convert = TRUE) %>%
  select(-temp1, -temp2) %>%
  arrange(i_track, time) %>%
  group_by(i_track) %>%
  # calculate time increments instead of total time:
  mutate(time = time - lag(time) %>% {.[1] = 0; .}) %>%
  ungroup() %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), NaN, .)) %>%
  mutate(name = ifelse(is.na(name), list(NULL), name)) %>%
  # small correction of transforming "NA" strings into real NA values:
  mutate(name = ifelse(name == "NA", NA, name)) %>%
  select(-parameterMetaSystem)

```

### Translate to pyramidi format


```{r }
dfm <- tab_measures(df_tuner_mod, ticks_per_beat)


dfm %>%
  split_midi_frame() %->% 
  c(df_meta, df_not_notes, df_notes_wide)

df_meta <-
  df_meta
```

## Export midi file from tuneR data

Now we have the data in 3 objects that `MidiFramer` can understand. First let's create an empty `MidiFramer` object:

```{r}
mfr <- MidiFramer$new()
```

We'll set the ticks per beat (see [here](https://urswilke.github.io/pyramidi/reference/MidiFramer.html#active-bindings))
```{r}
mfr$ticks_per_beat <- ticks_per_beat
```

First we'll pass the first two dataframes in the according fields of `mfr`:

```{r}
mfr$df_meta <- df_meta
mfr$df_not_notes <- df_not_notes
```

We'll add the last one with the method `MidiFramer$update_notes_wide()`, 
in order to pass the information to other fields 
(in particular `mfr$mf$midi_frame_nested`, 
which can be transformed back to a midi file; 
see `vignette("package_workflow")`):

```{r}
mfr$update_notes_wide(df_notes_wide)
```


```{r, eval=FALSE}
mfr$mf$write_file("round_tripped_midi_file.mid")
```

