### Pivot note dataframe to wide

Each note in the midi file is characterized by a `note_on` and a `note_off` event.
In order to generate a piano roll plot with ggplot2, we need to `pivot_wider()` those events.
This can be done with the function `widen_events()`:

```{r df_notes_wide}
df_not_notes <- 
  df_notes %>% 
  dplyr::filter(!stringr::str_detect(.data$type, "^note_o[nf]f?$")) 

df_notes_wide <-
  df_notes %>% 
  dplyr::filter(stringr::str_detect(.data$type, "^note_o[nf]f?$")) %>%
  # tab_measures(df_meta, df_notes, ticks_per_beat) %>%
  widen_events() %>%
  left_join(pyramidi::midi_defs)
df_notes_wide
```
In the new format, the data has half the number of rows.
The columns `m`, `b`, `t`, `ticks`, `time` and `velocity` are each replaced by
two columns with the suffix `_note_on` and `_note_off`.


### Plot the midi file information in a piano roll plot

Now we have the midi data in the right format for the piano roll plot:

```{r midi_piano_roll}
df_notes_wide %>%
  ggplot() +
  geom_segment(
    aes(
      x = m_note_on,
      y = note_name,
      xend = m_note_off,
      yend = note_name,
      color = velocity_note_on
    )
  ) +
  # each midi track is printed into its own facet:
  facet_wrap( ~ track,
              ncol = 1,
              scales = "free_y") + 
  guides(color=guide_colorbar(title="Note velocity")) +
  labs(title = "Piano roll of the note events in the midi file") +
  xlab("Measures") +
  scale_x_continuous(breaks = seq(0, 16, 4),
                     minor_breaks = 0:16) +
  scale_colour_gradient() +
  theme_minimal()

```
